Modelling BCG vaccination in the UK

### 2019-12-03 rmds/02-background.Rmd
added changes for background

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active TB in their lifetime.[@Houben:2016jp] Of this number around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, TB risk factors, TB treatment, global impact, and the impact TB in England and Wales are explored.

@@

TB has been associated with several risk factors, the most common of which is human immunodeficiency viruse (HIV). HIV increases the rate of activation by 20-fold and TB is the most common cause of AIDS-related death.[@Rottenberg2012] Increased risk of TB can also be the result of other medical conditions, such as diabetes, or lifestyle and environmental factors. These include smoking, low socioeconomic status, high density living, homelessness, incarceration, and drug use.[@Bhatti1995; @Narasimhan2013; @Story2007]

@@

1. Political commitment with increased and sustained financing
1. Case detection through quality-assured bacteriology
1. Standardized treatment with supervision and patient support
1. An effective drug supply and management system
1. Monitoring and evaluation system and impact measurement

@@

TB is a global disease with an estimated 10.4 million new cases in 2016,[@TheWorldHealthOrganization:2018va] of which 4.3 million were missed by health systems.[@WHO:2018]  Global incidence rates have decreased year-on-year since the early 2000s, with an average year-on-year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000s but this decline varies with region. Plot produced using getTBinR", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, viridis_end = 1) + theme(text = element_text(size = 12))
```

Regional incidence rates only tell part of the story as TB incidence rates vary significantly within regions. Six countries: India, Indonesia, China, Nigeria, Pakistan, and South Africa account for 60% of new cases. India, Indonesia, and Nigeria are thought to be responsible for nearly half of all undiagnosed cases.[@WHOTB2016] Figure \@ref(fig:map-global-tb) shows both regional similarities and countries, like Mongolia, that stand out as having higher TB incidence rates than surrounding countries.

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Incidence rates in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201", fig.height = 4}
map_tb_burden(verbose = FALSE) + 
  theme(text = element_text(size = 12))
```

@@

```{r global-regional-mortality-summary, fig.scap =  "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017.", fig.cap = "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017. Mortality rates from TB have been falling in all regions since 2000. Plot produced using getTBinR", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, metric = "e_mort_100k", 
                       denom = "e_inc_num", facet = "Area", legend = "none", 
                       scales = "free_y") +
  theme(text = element_text(size = 12))
```

@@

```{r map-global-tb-with-hiv, fig.scap = "Global map of estimated HIV in incidence TB (percent) in 2017.", fig.cap = "Global map of estimated HIV in incidence TB (percent) in 2017. Note the high percentage of TB cases with HIV in sub-saharan Africa. The percentages of HIV in incident TB in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.height = 4}
map_tb_burden(metric = "e_tbhiv_prct", verbose = FALSE) +
  theme(text = element_text(size = 12))
```

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all new TB cases globally in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 88% of rifampicin cases in Russia in 2017 also had MDR-TB, which is comparable to the global median of 83% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions Europe had the highest median percentage of rifampicin cases with MDR-TB (100%), with Africa having the lowest (67%). This variation may indicate areas in which drug resistance is developing (regions with low proportions of MDR to rifampicin resistance), from regions in which MDR-TB results from transmission or importation (regions with high proportions of MDR to rifampicin resistance).

```{r map-global-tb-with-rr, fig.scap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017.", fig.cap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017. Note that a far higher percentage of TB cases have rifampicin resistance in the former Soviet Union that in the rest of the world. The percentages of rifampicin resistances in incident TB in the legend refer to the lower bound for each colour. Map produced using getTBinR (see \\@ref(getTBinR) for details)", fig.height = 4}
map_tb_burden(metric = "e_rr_pct_new", verbose = FALSE) + 
  theme(text = element_text(size = 12))
```

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2017, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  arrange(desc(area)) %>% 
  mutate(`Median percentage of RR* cases with MDR** (95% IQR***)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi, digits = 1)) %>% 
  mutate(`Median Proportion of RR* cases with MDR** (95% IQR***)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median percentage of RR* cases with MDR** (95% IQR***)`)
                  )) %>% 
  select(Area = area, `Median percentage of RR* cases with MDR** (95% IQR***)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("* Rifampicin resistance", "** Multi-Drug Resistant TB", "*** Interquartile Range"), notation = "none") %>% 
  kable_styling(latex_options = "HOLD_position") %>% 
  {.}
```

@@

```{r plot-incidence, echo = FALSE, warning = FALSE, message = FALSE, fig.scap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB.", fig.cap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB. Interventions are highlighted with vertical lines, with linetype denoting the type of intervention, more information on each intervention is available in the corresponding table. Figure produced using tbinenglanddataclean", out.extra = ""}
plot_historic_tb_ew() + theme(legend.position = "top")+ 
  theme(text = element_text(size = 12))
```

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year-on-year since 2010 (when it was 49%).[@PHE2017] However, in 2016, 23.3% (420/1,800) of non-UK born cases had traveled outside the UK, with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

The majority of cases occur in urban areas. London alone accounts for 39% of cases, with an incidence rate of 25.1 (per 100,000, in 2016).[@PHE2017] England has few cases of MDR-TB cases, with only 68 cases recorded in 2016. Similarly the number of co-infections with HIV is low with only 3.8% of cases in 2015 having HIV - the majority of these cases were born in countries with high HIV prevalence. In 2016, 11.1% of TB cases in 2016 had at least one social risk factor, compared with 11.7% in 2015.[@PHE2017] In general cases with social risk factors are more likely to have drug resistant TB, worse TB outcomes, and to be lost to follow up. [@PHE2017] Amongst cases who were of working age in 2016, with a known occupation; 35.2% (1,491/4,240) were not in education or employment, 10.2% (432) were either studying or working in education; and 7.1% (304) were healthcare workers.[@PHE2017]

@@

Strain typing or whole genome sequencing is used to establish case clustering. This can be used to rule out transmission between cases but does not necessarily confirm transmission. Approximately 60% of cases cluster with at least one other case, and whilst this varies year-on-year, the fluctuations appear to be small (approximately 1-2%).[@PHE2017] Therefore interpreting any trend in TB transmission from the current strain typing data is difficult. Between 2010 and 2016, the median cluster size was 3 cases (range 2-244). In these clusters,  74.4% (2,141/2,878) consisted of less than 5 cases and only 8.8% of clusters had more than 10 cases [@PHE2017]. UK born cases were more likely to cluster than non-UK born cases (71.1%, 4,200/5,910 vs. 56.1%, 10,166/18,121).[@PHE2017]

@@

```{r plot-prop-pul, echo = FALSE, message = FALSE, warning = FALSE, fig.scap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB.", fig.cap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB. Figure produced using tbinenglanddataclean"}
plot_historic_prop_pul_tb() + theme(legend.position = "top") +
  theme(text = element_text(size = 12))
```

@@

The limited duration of protection has informed vaccination policy globally.[@Zwerling2011a] In countries where the BCG vaccination has been shown to be effective when given later in life, vaccination at school-age results in high levels of BCG effectiveness in young adults. As young adults are typically responsible for large amounts of TB transmission, this is likely to reduce TB incidence rates. Vaccination of neonates, on the other hand, provides protection against TB early in life. TB outcomes can be very poor at this time, but early life vaccination can lead to lower levels of protection later in life when transmission is more likely. This results in a trade-off with BCG vaccination of neonates being more effective in low effectiveness settings and in settings with lower TB incidence rates, whereas school-age vaccination is potentially more effective in settings with high BCG effectiveness and higher TB incidence rates.

@@

Until recently little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] There is some evidence that this reduction in all-cause mortality extends later in life.[@Rieckmann2016] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion (detection of TB bacterium in a sputum sample) .[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB[@Abubakar2013] mortality. The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with an RR of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-TB mycobacteria (NTM) infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va] This protection has been shown to greatly reduce the incidence of NTM lymph node inflammation and protect against Buruli ulcers for the first 12 months following vaccination (RR, 0.50 [95% CI, .37–.69]). A study also found that individuals with Buruli ulcers are less likely to develop osteomyelitis (bone infection) if they have a BCG scar (RR, 0.36 (95% CI: 0.22 to 0.58)).[@Zimmermann:2018io]

### Usage Globally

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100m doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The WHO recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. This includes universal vaccination of those at most risk of onwards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

As of 2011, among 180 countries with available data, 157 countries recommended universal BCG vaccination. The remaining 23 countries had either never implemented a universal programme or have switched to targeted vaccination of high risk individuals.[@Zwerling2011a] Most countries began universal programmes between the 1940s and 1980s due to high levels of TB incidence and strong evidence the effectiveness of the BCG vaccine.[@Hart1972] In the last 20 years 49 of these countries reported changing their vaccination programme with 27 countries reporting major changes in the last 10 years.[@Zwerling2011a] Globally, in countries that have BCG vaccination policies in place, coverage is estimated to be between 70% and 100%.

@@

In England, universal school-aged vaccination was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 78% in the white UK born population.[@Hart1972] In 2005, the UK shifted from this strategy to targeted vaccination in neonates deemed at high risk.[@Abubakar2013] This change was a reflection of current WHO vaccination policy,[@WHO2017] falling TB incidence rates, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] The impact of this change in policy is explored throughout this thesis but in particular in Chapter \@ref(sutherland), Chapter \@ref(direct-eff) and Chapter \@ref(model-results).

Since 2015, BCG vaccination has been included in the Cover Of Vaccination Evaluated Rapidly (COVER) programme, allowing coverage to be estimated in areas of England with universal vaccination (implemented due to high incidence rates and based on WHO guidelines). Coverage for areas in England implementing targeted vaccination remains unknown. In London current coverage estimates are made by Local Authority and range from 5.3% to 92.1%.[@PHE2017] These estimates may not be reliable as COVER has only recently begun to include returns for BCG, meaning that data quality may be poor. Prior to the switch to targeted neonatal vaccination coverage in those at school leaving age was thought to be approximately 75%.[@Sutherland1989]

>>>>

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active (symptomatic) TB in their lifetime.[@Houben:2016jp] Of those with active TB around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, TB risk factors, TB treatment, global impact, and the impact TB in England and Wales are explored.

@@

TB has been associated with several risk factors, the most common of which is human immunodeficiency viruse (HIV). HIV increases the rate of activation 20-fold and TB is the most common cause of AIDS-related death.[@Rottenberg2012] Increased risk of TB can also be the result of other medical conditions, such as diabetes, or lifestyle and environmental factors. These include smoking, low socioeconomic status, high density living, homelessness, incarceration, and drug use.[@Bhatti1995; @Narasimhan2013; @Story2007]

@@

1. political commitment with increased and sustained financing,
1. case detection through quality-assured bacteriology,
1. standardized treatment with supervision and patient support,
1. an effective drug supply and management system,
1. monitoring and evaluation system and impact measurement.

@@

TB is a global disease with an estimated 10.4 million new cases in 2016,[@TheWorldHealthOrganization:2018va] of which 4.3 million were estimated to have been missed by health systems.[@WHO:2018]  Global incidence rates have decreased year-on-year since the early 2000s, with an average year-on-year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary, with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean, incidence rates have remained relatively stable over the last 10 years.

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000s but this decline varies with region. Plot produced using `getTBinR`.", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, viridis_end = 1) + theme(text = element_text(size = 12))
```

Regional incidence rates only tell part of the story, as TB incidence rates vary significantly within regions. India, Indonesia, China, Nigeria, Pakistan, and South Africa account for 60% of new cases. India, Indonesia, and Nigeria are thought to be responsible for nearly half of all undiagnosed cases.[@WHOTB2016] Figure \@ref(fig:map-global-tb) shows both regional similarities and countries, like Mongolia, that stand out as having higher TB incidence rates than surrounding countries.

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Incidence rates in the legend refer to the lower bound for each colour. Map produced using `getTBinR`.", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201", fig.height = 4}
map_tb_burden(verbose = FALSE) + 
  theme(text = element_text(size = 12))
```

@@

```{r global-regional-mortality-summary, fig.scap =  "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017.", fig.cap = "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017. Mortality rates from TB have been falling in all regions since 2000. Plot produced using `getTBinR`.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, metric = "e_mort_100k", 
                       denom = "e_inc_num", facet = "Area", legend = "none", 
                       scales = "free_y") +
  theme(text = element_text(size = 12))
```

@@

```{r map-global-tb-with-hiv, fig.scap = "Global map of estimated HIV in incidence TB (percent) in 2018.", fig.cap = "Global map of estimated HIV in incidence TB (percent) in 2018. Note the high percentage of TB cases with HIV in sub-saharan Africa. The percentages of HIV in incident TB in the legend refer to the lower bound for each colour. Map produced using `getTBinR`.", fig.height = 4}
map_tb_burden(metric = "e_tbhiv_prct", verbose = FALSE) +
  theme(text = element_text(size = 12))
```

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all new TB cases globally in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 92% of rifampicin cases in Russia in 2018 also had MDR-TB, which is comparable to the global median of 81% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions, Europe had the highest median percentage of rifampicin cases with MDR-TB (87%), with the Western Pacific having the lowest (78%).

```{r map-global-tb-with-rr, fig.scap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2018.", fig.cap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2018. Note that a far higher percentage of TB cases have rifampicin resistance in the former Soviet Union that in the rest of the world. The percentages of rifampicin resistances in incident TB in the legend refer to the lower bound for each colour. Map produced using `getTBinR`.", fig.height = 4}
map_tb_burden(metric = "e_rr_pct_new", verbose = FALSE) + 
  theme(text = element_text(size = 12))
```

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2018, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  arrange(desc(area)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi, digits = 1)) %>% 
  mutate(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`)
                  )) %>% 
  select(Area = area, `Median Proportion of Rifampicin resistant cases with MDR* (2.5% to 97.5% quantiles)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("** Multi-Drug Resistant TB", "*** Only point estimates available"), notation = "none") %>% 
  kable_styling(font_size = 8, latex_options = "HOLD_position") %>% 
  {.}
```

@@

```{r plot-incidence, echo = FALSE, warning = FALSE, message = FALSE, fig.scap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB.", fig.cap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB. Interventions are highlighted with vertical lines, with linetype denoting the type of intervention, more information on each intervention is available in the corresponding table. Figure produced using `tbinenglanddataclean`.", out.extra = ""}
plot_historic_tb_ew() + theme(legend.position = "top")+ 
  theme(text = element_text(size = 12))
```

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year-on-year since 2010 (when it was 49%).[@PHE2017] 23.3% (420/1,800) of non-UK born cases had traveled outside the UK in 2016 with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

The majority of cases occur in urban areas. London alone accounts for 39% of cases, with an incidence rate of 25.1 (per 100,000, in 2016).[@PHE2017] England has few cases of MDR-TB cases, with only 68 cases recorded in 2016. Similarly the number of co-infections with HIV is low with only 3.8% of cases in 2015 having HIV - the majority of these cases were born in countries with high HIV prevalence. In 2016, 11.1% of TB cases in 2016 had at least one social risk factor, compared with 11.7% in 2015.[@PHE2017] In general cases with social risk factors are more likely to have drug resistant TB, worse TB outcomes, and to be lost to follow up.[@PHE2017] Amongst cases who were of working age in 2016, with a known occupation; 35.2% (1,491/4,240) were not in education or employment, 10.2% (432) were either studying or working in education; and 7.1% (304) were healthcare workers.[@PHE2017]

@@

Strain typing or whole genome sequencing is used to establish case clustering. This can be used to rule out transmission between cases, but does not necessarily confirm transmission. Approximately 60% of cases cluster with at least one other case, and whilst this varies year-on-year, the fluctuations appear to be small (approximately 1-2%).[@PHE2017] Therefore interpreting any trend in TB transmission from the current strain typing data is difficult. Between 2010 and 2016, the median cluster size was 3 cases (range 2-244). In these clusters,  74.4% (2,141/2,878) consisted of less than 5 cases and only 8.8% of clusters had more than 10 cases [@PHE2017]. UK born cases were more likely to cluster than non-UK born cases (71.1%, 4,200/5,910 vs. 56.1%, 10,166/18,121).[@PHE2017]

@@

```{r plot-prop-pul, echo = FALSE, message = FALSE, warning = FALSE, fig.scap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB.", fig.cap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB. Figure produced using `tbinenglanddataclean`."}
plot_historic_prop_pul_tb() + theme(legend.position = "top") +
  theme(text = element_text(size = 12))
```

@@

The limited duration of protection has informed vaccination policy globally.[@Zwerling2011a] In countries where the BCG vaccination has been shown to be effective when given later in life, vaccination at school-age results in high levels of BCG effectiveness in young adults. As young adults are typically responsible for large amounts of TB transmission, this is likely to reduce TB incidence rates. Vaccination of neonates, on the other hand, provides protection against TB early in life. TB outcomes can be very poor at this time, but early life vaccination can lead to lower levels of protection later in life when transmission is more likely. This results in a trade-off, with BCG vaccination of neonates being more effective in low effectiveness settings and in settings with lower TB incidence rates, whereas school-age vaccination is potentially more effective in settings with high BCG effectiveness and higher TB incidence rates.

@@

Until recently, little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] There is some evidence that this reduction in all-cause mortality extends later in life.[@Rieckmann2016] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion (detection of TB bacterium in a sputum sample) .[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB mortality.[@Abubakar2013] The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with a risk ratio (RR_ of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-TB mycobacteria (NTM) infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va] This protection has been shown to greatly reduce the incidence of NTM lymph node inflammation and protect against Buruli ulcers for the first 12 months following vaccination (RR, 0.50 [95% CI, .37–.69]). A study also found that individuals with Buruli ulcers are less likely to develop osteomyelitis (bone infection) if they have a BCG scar (RR, 0.36 (95% CI: 0.22 to 0.58)).[@Zimmermann:2018io]

### Usage globally

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100m doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The WHO recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. These include universal vaccination of those at most risk of onwards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

As of 2011, among 180 countries with available data, 157 countries recommended universal BCG vaccination. The remaining 23 countries had either never implemented a universal programme or have switched to targeted vaccination of high risk individuals.[@Zwerling2011a] Most countries began universal programmes between the 1940s and 1980s due to high levels of TB incidence and strong evidence of the effectiveness of the BCG vaccine.[@Hart1972] In the last 20 years 49 of these countries reported changing their vaccination programme with 27 countries reporting major changes in the last 10 years.[@Zwerling2011a] Globally, in countries that have BCG vaccination policies in place, coverage is estimated to be between 70% and 100%.

@@

In England, universal school-aged vaccination was introduced after an MRC trial in the 1950s estimated BCG's effectiveness at 78% in the white UK born population.[@Hart1972] In 2005, the UK shifted from this strategy to targeted vaccination in neonates deemed at high risk.[@Abubakar2013] This change was a reflection of current WHO vaccination policy,[@WHO2017] falling TB incidence rates, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] The impact of this change in policy is explored throughout this thesis but in particular in Chapter \@ref(sutherland), Chapter \@ref(direct-eff) and Chapter \@ref(model-results).

Since 2015, BCG vaccination has been included in the Cover Of Vaccination Evaluated Rapidly (COVER) programme, allowing coverage to be estimated in areas of England with universal vaccination (implemented due to high incidence rates and based on WHO guidelines). Coverage for areas in England implementing targeted vaccination remains unknown. In London, current coverage estimates are made by Local Authority and range from 5.3% to 92.1%.[@PHE2017] These estimates may not be reliable as COVER has only relatively recently begun to include returns for BCG, meaning that data quality may be poor. Prior to the switch to targeted neonatal vaccination, coverage in those at school leaving age was thought to be approximately 75%.[@Sutherland1989]
